Viscous Transport in Eroding Porous Media

### 2020-01-11 docs/transport/Response_to_Reviewer_1.tex

    \begin{quotation}
      \noindent
      Section 5 describes the numerical methods, with special attention
      paid to the quadrature method for computing the shear stress and
      the vorticity.
    \end{quotation}

    \begin{quotation}
      \noindent
      As a continuation of our previous work (Quaife \& Moore 2018), we
      have simulated dense suspensions and characterized transport in
      viscous eroding porous media. This is accomplished by using
      high-order time stepping methods and a combination of trapezoid
      and Barycentric quadrature methods to solve a BIE formulation of
      the Stokes equations. By using these numerical methods, we are
      able to perform stable simulations of erosion with $N = O(100)$
      discretization points, while the trapezoid rule would require
      $O(10^5)$ discretization points.
    \end{quotation}
\end{itemize}

\noindent
\comment{The transient increase of tortuosity is interesting, but the
explanation in Sec 6.4 is not satisfying. The authors proposed that the
increase of tortuosity can be explained by the sudden increase of
$x$-velocity when a grain vanishes. But in formula (4.4) or (4.5), the
$x$-velocity appears both in the numerator and the denominator, which
seems to me that the increase would just cancel. Then, as the authors
have pointed out, wouldn't the straighter streamlines resulted from the
vanishing of grains have decreased the tortuosity?}
\begin{itemize}
  \item This was an oversight on our behalf. The jump is actually caused
    by a small recirculation zone that vanishes when one of the eroding
    bodies vanishes. As the blue dots correspond to the Eulerian
    formulation, the sudden vanishing of this recirculation zone leads
    to a discrepancy between the two methods of calculating the local
    tortuosity.

  \item  While there is an error, at 7 of the porosity values around the
    jump in Figure 15a, the relative error between the Eulerian and
    Lagrangian formulations never exceeds $6 \times 10^{-3}$.

  \item We have updated the explanation to read:
    \begin{quotation}
      \noindent
      While it appears that the tortuosity does increase suddenly near
      the end of the simulation, this jump is due to a recirculation
      zone that vanishes as one of the grains completely erodes. This
      sharp increase is not present when the Lagrangian definition of
      the tortuosity is used. However, the relative difference between
      the tortuosities calculated with the Eulerian and Lagrangian
      formulations is less than $O(10^{-2})$.
    \end{quotation}
\end{itemize}

### 2020-01-11 docs/transport/erosion_transport.tex
finished up first round of revisions

The tortuosity can also be computed with an area integral. Assuming that
the flow is incompressible and not re-entrant, meaning that all
streamlines connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent
to~\citep{dud-koz-mat2011}
\begin{align}
  T = \left(\int_\Omega \|\uu(\yy)\|\, d\yy \right)
    \Bigg/
      \left(\int_\Omega u_1(\yy)\, d\yy \right),
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\citep{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in section~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes, represented by
the length of the red line segments, between neighboring grains. Two
grains are neighbors if they share an edge of the Delanuay triangulation
with nodes at the center of each eroding grain. (b) The pore sizes of an
eroded geometry. The black curve is the Weibull distribution with the
same first two moments as the data.}
\end{figure}
While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\citep{ali-par-wei-bre2017}, correlation
structures~\citep{leb-ded-dav-bou2007}, contaminant
transport~\citep{knu-car2005},
channelization~\citep{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
pore network models~\citep{bry-kin-mel1993, bry-mel-cad1993,
bij-blu2006}. To characterize the grain placement, we compute
distributions of pore sizes between neighboring grains.  To define
neighboring grains, we form the Delaunay triangulation using nodes
placed at the center of each eroding grain and at a collection of points
around the boundary of the porous media.  Then, we say that two grains
are neighbors if their centers share an edge of the
triangulation~\citep{dea-qua-bir-jua2018}. Finally, we define the pore
sizes to be the minimum distance between all pairs of neighboring
grains, and these distances are represented by the red line segments in
figure~\ref{fig:Eroding100gap}(a). We do not consider pores between
eroding bodies and the solid wall $\Gamma$, so some of the grains near
the porous region boundary only have two neighbors. Having defined the
pore sizes, we plot its distribution in
figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\citep{ioa-cha1993}. In section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.

@@

In figure~\ref{fig:Eroding100Transport}(a), we plot the tortuosity as a
function of the porosity.  The initial geometry has a porosity of $\phi
= 50.09\%$ and the tortuosity is $T = 1.20$.  The tortuosity is computed
with both the length of the streamlines~\eqref{eqn:tortuosity1} (red
stars) and using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in figure~\ref{fig:Eroding100tort}(c).
Again, the two tortuosity formulas give similar results.  For this
geometry, the tortuosity decreases monotonically at almost all
porosities.  However, the tortuosity undergoes a sudden increase near
the end of the simulation, and we have observed this behavior in other
examples.  The increase is caused by a single small body near the middle
of the channel being completely eroded.  While this results in
straighter streamlines, therefore reducing $\lambda$, the horizontal
flow, $u_1(y)$, increases since there is no longer a no-slip boundary,
and this increases the tortuosity.  We also compute the lines of best
fit using the porosity-tortuosity models~\eqref{eqn:tortuosityModels}.
Each of the models result in a root-mean-squared error ranging from
$4.60 \times 10^{-3}$ for model~\eqref{eqn:tortuosityModels4} to $7.20
\times 10^{-3}$ for model~\eqref{eqn:tortuosityModels3}. The black
dashed line in figure~\ref{fig:Eroding100Transport}(a) is the line of
best fit $\widehat{T}(\phi) = \phi^{-0.2459}$, with a root-mean-square
error of $5.50 \times 10^{-3}$.  We note a slightly better
root-mean-square error of $4.60 \times 10^{-3}$ is possible with the
model $\widehat{T}(\phi) = \left(1 + 0.1499(1-\phi)\right)^2$.

>>>>

The tortuosity can also be computed with an area integral. Assuming that
the flow is incompressible and not re-entrant, meaning that all
streamlines connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent
to~\citep{dud-koz-mat2011}
\begin{align}
  T = \left(\int_\Omega \|\uu(\yy)\|\, d\yy \right)
    \Bigg/
      \left(\int_\Omega u_1(\yy)\, d\yy \right),
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Our examples do contain recirculation zones, but they
are very small and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in section~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\citep{ali-par-wei-bre2017}, correlation
structures~\citep{leb-ded-dav-bou2007}, contaminant
transport~\citep{knu-car2005},
channelization~\citep{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
pore network models~\citep{bry-kin-mel1993, bry-mel-cad1993,
bij-blu2006}. To characterize the grain placement, we compute
distributions of pore sizes between neighboring grains.  To define
neighboring grains, we form the Delaunay triangulation using nodes
placed at the center of each eroding grain and at a collection of points
around the boundary of the porous media.  Then, we say that two grains
are neighbors if their centers share an edge of the
triangulation~\citep{dea-qua-bir-jua2018}. Finally, we define the pore
sizes to be the minimum distance between all pairs of neighboring
grains, and these distances are represented by the red line segments in
figure~\ref{fig:Eroding100gap}(a). We do not consider pores between
eroding bodies and the solid wall $\Gamma$, so some of the grains near
the porous region boundary only have two neighbors. Having defined the
pore sizes, we plot its distribution in
figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\citep{ioa-cha1993}. In section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.
\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes, represented by
the length of the red line segments, between neighboring grains. Two
grains are neighbors if they share an edge of the Delanuay triangulation
with nodes at the center of each eroding grain. (b) The pore sizes of an
eroded geometry. The black curve is the Weibull distribution with the
same first two moments as the data.}
\end{figure}

@@

In figure~\ref{fig:Eroding100Transport}(a), we plot the tortuosity as a
function of the porosity.  The initial geometry has a porosity of $\phi
= 50.09\%$ and the tortuosity is $T = 1.20$.  The tortuosity is computed
with both the length of the streamlines~\eqref{eqn:tortuosity1} (red
stars) and using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in figure~\ref{fig:Eroding100tort}(c).
Again, the two tortuosity formulas give similar results. While it
appears that the tortuosity does increase suddenly near the end of the
simulation, this jump is due to a recirculation zone that vanishes as
one of the grains completely erodes. This sharp increase is not present
when the Lagrangian definition of the tortuosity is used. However, the
relative difference between the tortuosities calculated with the
Eulerian and Lagrangian formulations is less than $O(10^{-2})$. We also
compute the lines of best fit using the porosity-tortuosity
models~\eqref{eqn:tortuosityModels}.  Each of the models result in a
root-mean-squared error ranging from $4.60 \times 10^{-3}$ for
model~\eqref{eqn:tortuosityModels4} to $7.20 \times 10^{-3}$ for
model~\eqref{eqn:tortuosityModels3}. The black dashed line in
figure~\ref{fig:Eroding100Transport}(a) is the line of best fit
$\widehat{T}(\phi) = \phi^{-0.2459}$, with a root-mean-square error of
$5.50 \times 10^{-3}$.  We note a slightly better root-mean-square error
of $4.60 \times 10^{-3}$ is possible with the model $\widehat{T}(\phi) =
\left(1 + 0.1499(1-\phi)\right)^2$.
